Stochastic Approximation Monte Carlo with Dynamic Update Factor (SAD)

### 2018-08-18 papers/histogram/sad.tex

While for SAMC the update factor is defined in the original
implementation, for SAD the update factor $\gamma_{t}^{\text{SA}}$ can
be thought of as $\nicefrac{\text{d}S}{\text{d}t}$. This tells us that
$t_0$ should have dimensions of entropy.  We argue that we can write
this entropy `term' geometrically as an area bounded by the $\ln{D(E)}$
i.e. the entropy. This gives the following
\begin{align}
\gamma_{t}^{\text{SAD}} =
\frac{N_E}{t}\frac{E_{H}-E_{L}}{3T_{\text{min}}}
\end{align}
where $N_E$ is the number of energies that have been found. The change
in the entropy the minimum temperature can be thought of as a slope of
the triangle made by connecting the minimum important energy and the
maximum entropy state.  The factor of $3$ comes from the ratio of areas
bounded by the $\ln{D(E)}$ and the minimum temperature line. The update
factor has the same $1/t$ behavior as the original SAMC algorithm;
however now every time a new energy is discovered during the course of
the simulation, the update factor experiences a jump. This behavior
allows SAD to dynamically prevent the update factor from decreasing too
rapidly. In contrast to SAD, Fig.~\ref{fig:gamma-vs-t}. shows the
monotonic decrease of the update factor for SAMC.  The update factor
for WL decreases like a step function while the update factor for SAD
dynamically fluctuates around a value less than 1 for early MC moves
(while finding new energy states).  Ultimately once all energy states
have been found, the update factor for SAD decreases in the same
monotonic fashion as SAMC.

{\color{red}
While for SAMC the update factor is defined in the original
implementation, for SAD the update factor $\gamma_{t}^{\text{SAD}}$ is
thought of as $\nicefrac{\text{d}S}{\text{d}t}$. This tells us that
$t_0$ should have dimensions of entropy.
We begin with an estimate of the average value of the entropy (relative
to the lowest entropy at $T_{\min}$).  If we assume a quadratic
dependence on energy, this is given by
\begin{align}
\langle S\rangle \approx \frac13 \frac{E({T=\infty}) - E(T_{\min})}{T_{\min}}
\end{align}
We approximate this energy difference by the energy difference
$E_H -E_L$ where these energies are defined below in the following
section.  Early in the simulation, when the energy range is still
expanding, this average entropy is a good choice
for $t_0$.  However, as the simulation progresses, our entropy factor
needs to scale with the total number of \davidsays{interesting} energy states that have been
found $N_S$, since the product $N_S\langle S\rangle$ corresponds to the
total change of entropy required, given that each of the $N_S$ states
will has a separate weight $w(E)$.  Finally, at \emph{long} times, after
all the energies have been found and the weights are being refined
we wish for a lower $t_0$ in order to reduce the fluctuations in the
weights that result from large values for $\gamma_t$.  We enable this
by tracking the last time we encountered a new energy $t_L$, and
gradually transition to lower $t_0$ (but still asymptotically scaling
as $\gamma_t \propto 1/t$ to ensure eventual convergence).  The SAD
expression for $\gamma_t$ which incorporates these ideas is:
\begin{align}
  \gamma_{t}^{\text{SAD}} =
     N_S \frac{
       \frac{E_{H}-E_{L}}{3T_{\text{min}}} + \frac{t}{t_L}
     }{
       N_S^2 + \frac{t^2}{t_L}
     }
\end{align}
This factor asymptotically has the same $1/t$ behavior as the original
SAMC algorithm;
however now every time a new energy is found in the
\davidsays{interesting energy range} during the course of
the simulation, the update factor experiences a jump. This behavior
allows SAD to dynamically prevent the update factor from decreasing too
rapidly. Figure~\ref{fig:gamma-vs-t} compares $\gamma_t$ for SAD, WL,
and SAMC.  The update factor
for WL remains high for many iterations, and then decreases very rapidly,
while the SAMC $\gamma_t$ remains constant before dropping as $1/t$.
The update factor for SAD in contrast fluctuates
dynamically around a value less than 1 for early MC moves
(while finding new energy states), before dropping in a \davidsays{similar
to SAD, but described after we see the plot}.
}

Since SAD does not explore \emph{all} energy states it needs to have a
reasonable idea
of what energy range corresponds to the temperature range of interest,
defined by $T_{\min}<T<\infty$.
The
simulation is responsible for determining and updating this energy
range.
Given the true entropy $S(E)$, we can define the interesting energy
range as
%\begin{align}
  $E(T_{\min}) <E< E(T=\infty)$
%\end{align}
where $E(T)$ is the energy that maximizes $S-E/T$.  In the course of the
simulation this precise energy may be challenging to evaluate accurately.
In order to ensure that we sample this entire energy range adequately,
we define two energy limits:  a high energy $E_H$ and a low
energy $E_L$ to define the range over which the energy histogram
is made flat. $E_H$ is chosen to be the greatest energy that
maximized the entropy at any time during the simulation.  Similarly,
$E_L$ is the lowest energy that ever maximized $S(E)-E/T_{\min}$ during
the simulation.  This creates a ``ratcheting'' effect, in which $E_H$
may only increase, while $E_L$ may only decrease over the course of the
simulation, which results in a conservative estimate of the range of
energies that need be sampled.

Outside of the energy range $E_L \le E \le E_H$, the weights are
determined from the accumulated energy histogram. For each energy $E_i
> E_H$, the energy weights are calculated
\begin{align}
w(E_i)_{t+1} = w(E_H)_{t} + \ln\left(\frac{H(E_i)}{H_{\max}}\right)
\end{align}
assuming of course that the simulation has collected statistics at that energy.
For each energy $E_i
< E_L$, the energy weights are calculated
\begin{align}
w(E_i)_{t+1} = w(E_L)_{t} + \ln\left(\frac{H(E_i)}{H_{\max}}\right) +
\frac{E_i - E_L}{T_{\min}}
\end{align}

>>>>

While for SAMC the update factor is defined in the original
implementation, for SAD the update factor $\gamma_{t}^{\text{SAD}}$ is
thought of as $\nicefrac{\text{d}S}{\text{d}t}$. This tells us that
the SAMC parameter
$t_0$ should have dimensions of entropy.
We begin with an estimate of the average value of the entropy (relative
to the lowest entropy at $T_{\min}$).  If we assume a quadratic
dependence on energy, this is given by
\begin{align}
\langle S\rangle \approx \frac13 \frac{E({T=\infty}) - E(T_{\min})}{T_{\min}}
\end{align}
We approximate this energy difference by
$E_H -E_L$ where $E_H$ and $E_L$ are defined below.
The entropy numerator of the update factor
in general should scale with the total number of interesting
energy states $N_S$, since updates to the weights are distributed between
that many energy states.  The product $N_S\langle S\rangle$ is the
total change of entropy required (starting from constant weights) to find
the true entropy, and puts a lower bound on the convergence time.
After \emph{long} times, when
all the energies have been long ago found,
we wish for a lower $t_0$ in order to more rapidly refine the remaining
error in entropy.  We enable this
by tracking the last time we encountered a new energy $t_L$, and
gradually transition to lower $t_0$ (but still asymptotically scaling
as $\gamma_t \propto 1/t$ to ensure eventual convergence).  Finally,
we wish for an update factor that is \emph{never} greater than 1, because
a very large update factor could introduce errors in entropy that may take
many iterations to remove.  The SAD
expression for $\gamma_t$ which incorporates these ideas is:
\begin{align}
  \gamma_{t}^{\text{SAD}} =
     \frac{
       \frac{E_{H}-E_{L}}{3T_{\text{min}}} + \frac{t}{t_L}
     }{
       \frac{E_{H}-E_{L}}{3T_{\text{min}}} + \frac{t}{N_S}\frac{t}{t_L}
     }
\end{align}
This factor asymptotically has the same $1/t$ behavior as the original
SAMC algorithm;
however now every time a new energy is found to be important
during the course of
the simulation, the update factor will experience a jump. This behavior
allows SAD to dynamically prevent the update factor from decreasing too
rapidly.

Figure~\ref{fig:gamma-vs-t} compares $\gamma_t$ for SAD, WL,
and SAMC.  The update factor
for WL remains high for many iterations, and then decreases very rapidly,
while the SAMC $\gamma_t$ remains constant before dropping as $1/t$.
The update factor for SAD in contrast fluctuates
dynamically around a value less than 1 for early MC moves
(while finding new energy states), before dropping in a \davidsays{similar
to SAD, but described after we see the plot}.

Since SAD does not explore \emph{all} energy states it needs determine
what energy range corresponds to the temperature range of interest,
defined by $T_{\min}<T<\infty$.
The
simulation is responsible for determining and updating this energy
range.
Given the true entropy $S(E)$, we can define the interesting energy
range as
%\begin{align}
  $E(T_{\min}) <E< E(T=\infty)$
%\end{align}
where $E(T)$ is the energy that maximizes $S-E/T$.  In the course of the
simulation this precise energy is challenging to evaluate accurately.
In order to ensure that we sample this entire energy range adequately,
we define two energy limits:  a high energy $E_H$ and a low
energy $E_L$, which define the range over which the energy histogram
is made flat. At move $t$, $E_H$ and $E_L$ are the greatest and lowest
energy that has prior to that move
had the highest histogram value (i.e. been visited the most times) in
the course of the simulation.
This definition results in a ``ratcheting'' effect, in which $E_H$
may only increase, while $E_L$ may only decrease over the course of the
simulation, which results in a conservative estimate of the range of
energies that need be sampled.

During the simulation when considering a move outside of the energy
range $E_L \le E \le E_H$, Boltzmann weights are used.  If $E\ge E_H$,
the weight is taken to be
\begin{align}\label{eq:highw}
  w(E>E_H) &= w(E_H),
\end{align}
which corresponds to an infinite temperature.
This choice ensures that
if the maximum in entropy is at an energy $E_{\max}>E_H$, then the energy
$E_{\max}$
will eventually have the highest number of counts and the ratcheting will
result in $E_H\ge E_{\max}$ .
At lower energies, Boltzmann weights corresponding to the minimum temperature
are used:
\begin{align}\label{eq:loww}
  w(E<E_L) &= w(E_L)e^{-\frac{E_L-E}{T_{\min}}}.
\end{align}
This choice has the result that if the energy $E_{\min}$ at which the free
energy at $T_{\min}$ is minimized is less than $E_L$, the lower energy
limit will ratchet down to include $E_{\min}$.
Each time we change the value of $E_H$ or $E_L$, the weights within the
new portion of the interesting energy range are set to the expressions
in Equations~\ref{eq:highw} and~\ref{eq:loww}.
